---
title: "EDA and pre-processing of ships data"
author: "Miguel Eersel"
date: "14 May 2021"
output: 
  html_document:
    code_folding: show
    toc: TRUE
    toc_float:
      collapsed: FALSE 
    theme: cerulean
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
   
# Core packages -----
library(tidyverse)
library(lubridate)

# data import
library(vroom)

# data manipulation
library(data.table)
library(tidyfast)

# Helper packages
library(fs)
library(janitor)
library(skimr)
library(DataExplorer)
library(here)

# BSU packages
library(tidyquant)
library(timetk)

# geospatial
library(geodist)
library(sf)

# tables
library(gt)
library(reactable)
library(reactablefmtr)


```

```{r function-get-distance}

get_distance <-  function(df = df){
  
# Purpose: PAIRWISE calculation of distance
  
# The time-distance path of a ship is retrieved via the lon and lat variables.

# INPUT: a dataframe of the travel route with datetime lon, lat observation per ship
  
  
# Step 1: sort datetime in ascending order
  df <-  df %>% 
          arrange(datetime) %>% 
# Step 2: retrieve tehe previous lon/lot geospatial datetime position    
          mutate(lag_lat = lag(lat, 1),
                 lag_lon = lag(lon, 1))
          
          
          
          
df2 <-  df %>% 
          drop_na(lag_lat)
  
lat1 <-  df2 %>% 
          pull(lat)

lat2 <-  df2 %>% 
          pull(lag_lat)

lon1 <-  df2 %>% 
          pull(lon)

lon2 <-  df2 %>% 
          pull(lag_lon)

# Step 3: calculate pairwise distances
# distance between pairs {(lon1, lat1) and (lon2, lat2)}

# result is stored in the dist_vec
dist_vec <- geodist::geodist_vec(
                     x1 = lon1,y1 = lat1,
                     x2 = lon2, y2 = lat2,
                     paired = TRUE, 
                     measure = "haversine")

# due to the lag() operation the first position is NA,. Replace NA with ZERO
dist_vec <-  c(0,dist_vec)  
  

  
  df_result <- df %>% 
        mutate(
           lag_datetime = dplyr::lag(datetime, n = 1),

           # Calculate the   time difference between two (2) consecutive observations in unit hours          
           traveltime_hours = lubridate::interval(start = lag_datetime,
                                                  end = datetime) / lubridate::dhours(x = 1)
        ) %>% 
    
       mutate(
               diff_lat = (lat - lag_lat) %>%  round(5),
               diff_lon  = (lon - lag_lon) %>%  round(5),

              

              consecutive_dist_m = round(dist_vec,digits = 0), # round distance to whole meters
              
              speed_meters_per_hour = consecutive_dist_m / traveltime_hours,
               
              speed_knots = (speed_meters_per_hour / 1852)
              ) 
    
# return the result
  return(df_result)
}

```

```{r function-get-longest-distance}

get_longest_distance <- function(df = df){

# PURPOSE: Retrieve the longest path between two(2) consecutive observations 

# INput: dataframe df with the distance between each consecutive observation (per ship)  
    
  df_calc <- df %>% 
    
# select longest distance    
    slice_max(order_by = consecutive_dist_m,n = 1) %>% 
    
# select the most recent observation    
    arrange(desc(datetime)) %>% 
    slice(1)
  
  return(df_calc)
  
}


```

```{r read-ships-data}

file_name_path <- here::here("00_data","ships.csv")


# read the ship data and store it into ship_raw_tbl
ships_raw_tbl <- vroom::vroom(file = file_name_path,
                              col_types = cols()) %>% 
                 janitor::clean_names() %>% 
                 mutate(ship_id = as.character(ship_id),
                        ship_id = stringr::str_squish(ship_id),
                        shiptype = as.character(shiptype),
                        lat = round(lat,5),
                        lon = round(lon,5)
                        ) %>% 
                  
                rename(ship_type_id = shiptype)

```

<br>

------------------------------------------------------------------------

## Data Cleaning steps

The **ship_id** is a **unique** identifier for a ship name. A test is performed if this in indeed true. The findings are the shipname + shipptype combination is NOT unique.

The following ship_id are NOT unique in combination with **shipname** and **ship_type**.

------------------------------------------------------------------------

```{r data-cleaning-steps}

ships_raw_tbl %>% 
  count(ship_id, shipname, ship_type, name="nr_obs") %>%  #  retrieve the unique combinations of ship_id and ship_names
  add_count(ship_id, name = "ship_id_counts") %>%  # Count how many time a ship_id occurs. For a unique ship_id a ship_id value may only occur once.. Aad the counted value, n , as a new coloumn
  filter(ship_id_counts > 1) %>% 
  arrange(desc(ship_id_counts), ship_id,desc(nr_obs), shipname,ship_type) %>% 
  gt::gt(groupname_col = "ship_id",
         rownames_to_stub = TRUE,rowname_col = "shipname") %>% 
  
  
  
  gt::tab_style(style = list(cell_fill(color = "lightcyan")),
                locations = cells_row_groups()) %>% 
  
  gt::tab_options(table.width = pct(100),
                  table.font.size = px(11),
                  column_labels.font.size = px(15), 
                  column_labels.font.weight = "bold",) %>% 
  
  
  gt::opt_row_striping()






```

------------------------------------------------------------------------

For data cleaning of these non-unique ship_id, the website of [MarineTraffic AIS signal](https://www.marinetraffic.com/en/ais/home/centerx:-12.0/centery:25.0/zoom:4) has been used to investigate which **ship_id** corresponds with which **shipname** and **ship_type**.

According to the information retrieved from this website the following cleaning of the non-unique ship_ids will be performed:

| ship_id | shipname          | ship_type   |
|---------|-------------------|-------------|
| 4666609 | BLACKPEARL        | Unspecified |
| 315731  | ODYS              | Tug         |
| 315950  | WLA-311           | Fishing     |
| 316404  | KAPITAN BORCHARDT | Pleasure    |
| 316482  | WYSPA SZCZESCIA   | Pleasure    |
| 345254  | IVAN KRUZENSHTERN | Tug         |
| 347195  | GAZPROMNEFT WEST  | Tanker      |
| 364937  | BOMAR MOON        | Cargo       |
| 3653787 | ARGO              | Tug         |
| 406999  | SEASTAR ENDURANCE | Cargo       |
| 4190565 | OCEANOGRAF        | Tug         |
| 757619  | PRINCE OF WAVES   | Cargo       |

<br>

For some ships , mainly of the tug boat type, no shipname was available. For these ships , the text **SAT-AIS** was registered as the shipname. In the data cleaning/data imputation steps the SAT-IS text was replaced by the unique SAT-IAS code of these ships.

------------------------------------------------------------------------

### **Calculation logic**

As the **Shiny demo app** is an app with limited interactive calculations I have chosen to **pre-process** the raw (csv) data into its **semi-final data form** for the Shiny Dashboard.

The main calculation logic is performed in this RMarkdown file which is a kind a **ETL** data pre-processor.

Pairwise distance calculation\
- function: get_distance \<- function(df = df)

Longest distance between two consecutive observation\
- function: get_longest_distance \<- function(df = df)

------------------------------------------------------------------------

### **Nested dataframe**

For the purpose of code development and readability, **nested** data frames have been used to keep the (large) dataframe (approx. 3 million records with 1210 shipid) organized.

**Nested** dataframes go hand-in-hand with the *map()* function.\
All calculations per ship are performed via the *map()* function in combination with a *nested* dataframe.

------------------------------------------------------------------------

### **Distance calculation**

A ***pairwise*** geo-spatial distance calculation method has been selected from the **{geodist}** package.

Speed performance research\
First the distance calculation function **st_distance()** from the **{sf}** package had been selected but the (speed) performance was not good. The **geodist_vec()** function of the {geodist} package is at least 5x times faster.

------------------------------------------------------------------------

### **Results**

The results of the distance and other calculations have been stored in **rds files**. The input and output data files are stored in the **00_data** sub-folder.

The rds files are the input data files for the Shiny Ships Dashboard application.

------------------------------------------------------------------------

### Shiny application

The Shiny application is on `shinapps.io`

\- Link to the app [AIS Ship Dashboard](https://miguel-eersel.shinyapps.io/shipdashboard)

### Github repository

The code of the app as well as this RMakrdown file can be found on my Public Github repository.

\- Link to the Github repositoty [GitHub AIS Ship Dashboard](https://github.com/meersel/shipdashboard)

-   The rds files as too larges to be uploaded to Github.

    -   The rds output files can be created with this RMarkdown (ETL) file.

```{r create-nested-ship-raw-tbl}

ship_nested_raw_tbl <- ships_raw_tbl %>% 
  group_nest(ship_id, shipname, ship_type)


```

```{r create-ship-nested-clean-tbl}

ship_nested_clean_tbl <- ship_nested_raw_tbl %>% 

  mutate(
    shipname = case_when(
    ship_id == "4666609" ~ "BLACKPEARL",
    ship_id == "315731" ~ "ODYS",
    ship_id == "315930" ~ "WLA-311",
    ship_id == "315950" ~ "WLA-311",
    ship_id == "316404" ~ "KAPITAN BORCHARDT",
    ship_id == "347195" ~ "GAZPROMNEFT WEST",
    ship_id == "364937" ~ "BOMAR MOON",
    ship_id == "3653787" ~ "ARGO",
    ship_id == "406999" ~ "SEASTAR ENDURANCE",
    ship_id == "757619" ~ "PRINCE OF WAVES",
    ship_id == "316482" ~ "WYSPA SZCZESCIA",
    ship_id == "345254" ~ "IVAN KRUZENSHTERN",
    
    TRUE ~ shipname),

# Ship OCEANOGRAF has two ship types. The "Tug" ship type is selected.        
    ship_type = case_when(
      ship_id == "4190565" ~ "Tug",
      TRUE ~ ship_type),


  ) %>%
  
  mutate(
# For ships with the [SAS-AIS] as their ship name. the ship name will be replaced by the [SAS-AIS] number = ship_id        
   shipname = case_when(
     stringr::str_detect(shipname,"\\[SAT-AIS\\]") ~ ship_id,
     TRUE ~ shipname),
   shipname = case_when(
     stringr::str_detect(shipname,"SAT AIS") ~ ship_id,
     TRUE ~ shipname),
   
   
   ) %>% 
  
  unnest(cols = data) %>%  
  add_count(ship_id, shipname, ship_type, name = "nr_obs") %>% 
  group_nest(ship_id,shipname, ship_type, nr_obs) 


# add ship_id  to shapname in case that shipname+ship_type combination is not unique
ship_nested_clean_tbl <- ship_nested_clean_tbl %>% 
  add_count(shipname, ship_type, sort = TRUE) %>% 
  arrange(desc(n), shipname) %>% 
  mutate(shipname = case_when(
    n > 1 ~ paste0(shipname,"-",ship_id),
    TRUE ~ shipname
  )) %>% 
  select(-n)
    


 write_rds(x = ship_nested_clean_tbl,file = "00_data/ship_nested_clean_tbl.rds")

 
 
ship_clean_tbl <- ship_nested_clean_tbl %>% 
  unnest(data)

 write_rds(x = ship_clean_tbl,file = "00_data/ship_clean_tbl.rds")
 

```

```{r calculate-distance-between-geopoints}

ship_dist_tbl <-  ship_clean_tbl %>%
   select(starts_with("ship"), lon, lat, speed, datetime, nr_obs) %>%
  
  group_nest(ship_id) %>% 
#  filter(ship_id %in% c("1036966", "1037103")) %>% 
    mutate(dist_nested_lst = purrr::map(.x = data,
                                  .f = get_distance)) %>%
  ungroup() %>% 
  select(-data) %>% 
  unnest(cols = dist_nested_lst)

ship_dist_nested_tbl <-  ship_dist_tbl %>% 
                          group_nest(ship_id, shipname, ship_type, nr_obs)
  
# store data a a nested dataframe
write_rds(x = ship_dist_nested_tbl,file = "00_data/ship_dist_nested_tbl.rds")


```

```{r calculate-the-longest-distance-per-shipid}

# Calculate the longest distance using teh fuction get_longest_distance()

ship_longest_dist_tbl <- ship_dist_nested_tbl %>% 
  mutate(longest_distance = map(data, .f=get_longest_distance)) %>% 
  select(-data) %>% 
  unnest(cols = longest_distance) %>% 
  relocate(shipname, consecutive_dist_m)


write_rds(ship_longest_dist_tbl,file = "00_data/ship_longest_dist_tbl.rds")

```

<br>

------------------------------------------------------------------------

### Basic Ship type analysis (counting)

-   Dynamic sortable table **{reactable}**\
-   Basic statistics about the **ship type**

------------------------------------------------------------------------

\`

```{r dynamic-table-shiptype}



 data <- ship_nested_clean_tbl %>% 
  select(-data) %>% 
  group_by(ship_type) %>% 
   summarise(
     tot_types = n(),
     tot_obs   = sum(nr_obs)
  ) %>% 
  ungroup() %>% 
  mutate(avg_obs = (tot_obs/tot_types) %>%  round(0))
   
  reactable::reactable(data = data,
    columns = list(
      avg_obs = colDef(align = "left",
                       cell = reactablefmtr::data_bars(data,
                                                       colors = "red",
                                                       number_fmt = scales::comma_format(accuracy = 1))),
      tot_types = colDef(align = "left",
                       cell = reactablefmtr::data_bars(data,
                                                       colors = "darkgreen",
                                                       number_fmt = scales::comma_format(accuracy = 1))),
    
      tot_obs = colDef(align = "left",
                       cell = reactablefmtr::data_bars(data, 
                                                       colors = "dodgerblue",
                                                       number_fmt = scales::comma_format(accuracy = 1)))
    ),
    sortable = TRUE
    ) 
  
  




```

```{r dummy-chunck-for-testing, eval = FALSE}


```

<br>

------------------------------------------------------------------------

### Contact information:

Author: Miguel Eersel

e-mail: m.eersel\@outlook.com

mobiel: +597-864 2328 (Suriname, GMT - 3 hours)
